Parsing Postfix log files

### 2007-11-12 doc/logparser.tex
More corrections.

Some mail clients appear to send the following sequence of commands during
the \SMTP{} session:

@@

Once again Postfix does not log a message making the client's behaviour
clear, so once again heuristics are required to identify when this
behaviour occurs.  In this case a list of all mails accepted during a
connection are saved in the connection state, and the accepted mails are
examined when the disconnection line is parsed and the disconnection action
executed.  Each mail is checked for the following characteristics:

@@

    \item Were there exactly two \daemon{smtpd} log lines for the mail?
        There should be a connection line and a mail accepted line.

@@

If both checks are successful then the mail is assumed to be one of the
offending bogus mails and is discarded.  There will be no further entries
logged for such mails, so without identifying and discarding them they
accumulate in the state table and will cause clashes if the queueid is
reused.  The mail cannot be entered in the database as the only data
available is the client hostname and IP address, but the database schema
requires many more fields be populated (see section~\refwithpage{database
schema}).

@@

The DATA phase of the \SMTP{} conversation is where the headers and body of
the mail are transferred.  Sometimes there is a timeout or the connection
is lost\footnote{For brevity's sake timeout will be used throughout this
section, but everything applies equally to lost connections.} during the
DATA phase; when this occurs Postfix will discard the mail and the parser
needs to discard the data associated with that mail.  It seems more
intuitive to save the data to the database, but if a timeout occurs there
won't yet have been any data gathered for the mail, so there is none
available to save; the timeout is saved with the connection data instead.

@@

    \item Because of ESMTP pipelining the client can send MAIL FROM, RCPT
        TO and DATA commands in one packet, instead of sending each command
        individually and waiting for the reply before sending subsequent
        commands.  If pipelining is used and Postfix rejects the sender
        address or any of the recipient addresses there is no way for the
        client to tell which command was rejected.  Some clients make no
        attempt to recover and disconnect uncleanly.

@@

In the case where the \daemon{cleanup} line is still pending the algorithm
updates a global cache of queueids, adding the queueid and the timestamp
from the timeout line.  When the next \daemon{cleanup} line is parsed for
that queueid the cache will be checked, and the line will be deemed part of
the discarded mail and discarded if it meets the following requirements:

@@

The next \daemon{cleanup} line must meet the criteria above for it to be
discarded because not every connection where a timeout occurs will have a
\daemon{cleanup} line logged for it; if the algorithm blindly discarded the
next \daemon{cleanup} line it would in some cases be mistaken.  Whether or
not the next \daemon{cleanup} line is discarded the queueid will be removed
from the cache of timeout queueids when the next \daemon{pickup} line
containing that queueid is parsed.

During the TIMEOUT action if the mail lacks a \daemon{cleanup} line, the
queueid and other data about the mail, particularly the timestamp of the
last log line, are added to a global cache of timeout queueids.  This cache
is subsequently consulted during the MAIL\_PICKED\_FOR\_DELIVERY action,
and if the queueid exists and the \daemon{cleanup} line passes the checks
listed above it will be discarded.

@@

Occasionally the \daemon{pickup} line logged when mail is submitted locally
via sendmail appears later in the logfile than the \daemon{cleanup} line
for that mail.  This seems to occur during periods of particularly heavy
load, so is most likely due to process scheduling vagaries.  Normally if
the queueid given in the \daemon{pickup} line exists a warning is generated
by the \daemon{pickup} action, but if the following conditions are met it
is assumed that the lines are out of order:

@@

    \item There is less than a five second difference in the timestamps of
        the \daemon{cleanup} and \daemon{pickup} lines.

@@

As always with heuristics there may be circumstances in which these
heuristics incorrectly match, but none have been identified so far.

@@

    \item Postfix is stopped or its configuration is reloaded.  When this
        happens all \daemon{smtpd} processes exit, and all entries in the
        connections state table must be committed and deleted from the
        state table.  The queueid state table is untouched.

    \item Postfix sometimes forcibly kills a \daemon{smtpd}, so the active
        connection must be committed and deleted from the connections state
        table.

    \item Occasionally a \daemon{smtpd} will die or exit uncleanly, so the
        active connection must be committed and deleted from the
        connections state table.

@@

The above appear to cover all situations where a \daemon{smtpd} suddenly
stops logging.  In addition to removing an active connection the last
accepted mail may need to be discarded, as detailed in
section~\refwithpage{timeouts-during-data-phase}.

@@

Occasionally a log file will have out of order log lines which cannot be
dealt with by the techniques described in sections~\refwithpage{tracking
re-injected mail},~\refwithpage{discarding cleanup
lines}~and~\refwithpage{pickup logging after cleanup}.  In the 93 log files
used for testing this occurs only five times in 60,721,709 log lines, but
for completeness of the algorithm it should be dealt with.  The five
occurrences in the test log files have the same characteristics: the
\daemon{local} log line showing delivery to a local mailbox occurs after
the \daemon{qmgr} log line showing removal of the mail from the queue
because delivery is completed.  This causes problems: the mail is not
complete, so entry into the database fails; a new mail is created when the
\daemon{local} line is parsed and remains in the state tables; four
warnings are issued per pair of out of order log lines.

The solution to this problem is to examine the list of programs which have
logged messages for each mail, comparing the list against a table of
known-good combinations of programs (this check is performed during the
COMMIT action).  If the mail's combination is found in the valid list the
mail can be entered in the database; if the combination is not found entry
must be postponed and the mail flagged.  The SAVE\_BY\_QUEUEID action
checks for the flag and retries entry if it's found; if the additional log
lines have caused the mail to reach a valid combination entry will succeed,
otherwise it must be postponed once more.

The list of valid combinations is given below.  Every mail will
additionally have log entries from \daemon{cleanup} and \daemon{qmgr}; any
mail may also have log entries from \daemon{bounce}, \daemon{postsuper}, or
both.

@@

    \item [\daemon{local}:] Local delivery of bounce notification, or local
        delivery of forwarded/tracked mail.

@@

    \item [\daemon{local}, \daemon{smtpd}:] Mail accepted from a remote
        client, local delivery.

    \item [\daemon{pickup}, \daemon{smtp}:] Locally submitted mail, remote
        delivery.

    \item [\daemon{smtp}:] Remote delivery of forwarded/tracked mail.

@@

This applies to accepted mails only, not to rejected mails.

@@

This section has presented the parsing algorithm, divided into: a high
level overview of the basic algorithm, followed by the first group of
difficulties thwarted.  The revised algorithm was detailed, preceded by a
flow chart depicting the paths a mail may take through the algorithm.  The
actions which are available to rules were described in full, and finally
the remaining complications discovered and overcome in the process of
refining and completing the algorithm.  Detecting, diagnosing and defeating
complications forms the largest single subsection of this document,
mirroring the development of the parser,  The complications are described
in the order they were overcome, with subsequent problems affecting fewer
mails (often by an order or magnitude), though the time required to solve
problems increased with each successive problem.

@@

Coverage of mails is much more difficult to determine accurately than
coverage of lines.  The parser can dump its state tables in a human
readable form; examining these tables with reference to the log files is
the best way to detect mails which were not handled properly (many of the
complications discussed in section~\refwithpage{additional complications}
were detected in this way).  The parser issues warnings when it detects any
errors, some of which may alert the user to a problem, e.g.\ when a queueid
is reused before the previous mail is fully dealt with, when a queueid or
\pid{} is not found,\footnote{There will often be warnings about a missing
queueid or \pid{} in the first few hundred or thousand log lines because the
earlier log lines for those connections or mails are in the previous log
file; loading the saved state from the previous log file will solve this
problem.} or when there are problems tracking a child mail (see
section~\refwithpage{tracking re-injected mail}).  There should be few or
no warnings, and the state table should only contain entries for mails
which had yet to be delivered when the logfiles ended, or started before
the logfiles began.

@@

The dips at logfiles 22 and 62--68 correspond to peaks in log file size in
graph~\refwithpage{execution time vs file size vs number of lines graph},
and peaks in graph~\refwithpage{execution time vs file size vs number lines
factor} (where a peak means that more lines are processed per second, i.e.\
performance is better).  The explanation for this took some time to arrive
at, but it turns out to be reasonably simple.  The large log files in
question were caused by a mail forwarding loop, where the distribution of
log lines is quite different to normal, resulting in different performance
characteristics.

@@

% The contents of the glossary.
\glossary{name={ESMTP},description={
    ESMTP is Extended \SMTP{}, defined in \RFC{}~1869~\cite{RFC1869}.
}}

\glossary{name={SQLite3},description={
    \textit{SQLite is a small C library that implements a self-contained,
    embeddable, zero-configuration SQL database engine.\/} SQLite3 is an
    \SQL{} implementation focusing on correctness, simplicity and speed.
    Unlike other \SQL{} implementations it does not require a separate
    server process, greatly simplifying deployment of an application
    utilising it.  More details can be found at~\cite{sqlite-features} or
    \url{http://www.sqlite.org/}.
}}

@@

>>>>

Some mail clients appear\footnote{Due to privacy concerns no attempt has
been made to identify either the users or the software which exhibits this
behaviour.} to send the following sequence of commands during
the \SMTP{} session:

@@

Once again Postfix does not log a message making the client's behaviour
clear, so once again heuristics are required to identify when this
behaviour occurs.  In this case a list of all mails accepted during a
connection is saved in the connection state, and the accepted mails are
examined when the disconnection action is executed.  Each mail is checked
for the following characteristics:

@@

    \item Were there exactly two \daemon{smtpd} log lines for the mail?
        There should be a connection line and a mail accepted line, nothing
        else.

@@

If both checks are successful then the mail is assumed to be one of the
offending bogus mails and is discarded.  There will be no further entries
logged for such mails, so without identifying and discarding them they
accumulate in the state table and will cause clashes if the queueid is
reused.  The mail cannot be entered in the database as the only data
available is the client hostname and IP address, but the database schema
requires many more fields be populated (see
sections~\refwithpage{connections table} and~\refwithpage{results table}).

@@

The DATA phase of the \SMTP{} conversation is where the headers and body of
the mail are transferred.  Sometimes there is a timeout or the connection
is lost\footnote{For brevity's sake timeout will be used throughout this
section, but everything applies equally to lost connections.} during the
DATA phase; when this occurs Postfix will discard the mail and the parser
needs to discard the data associated with that mail.  It seems more
intuitive to save the data to the database, but if a timeout occurs there
won't yet have been any data gathered for the mail, so there is none
available to save; the timeout is recorded with the connection data
instead.

@@

    \item \ESMTP{} pipelining allows the client to send MAIL FROM, RCPT TO
        and DATA commands in one packet, instead of sending each command
        individually and waiting for the reply before sending subsequent
        commands.  If pipelining is used and Postfix rejects the sender
        address or any of the recipient addresses there is no way for the
        client to tell which command was rejected.  Some clients make no
        attempt to recover and disconnect uncleanly; others presumably
        stop pipelining and restart the mail sending operation.  The client
        could attempt to parse the error message, but that would be
        extremely error prone, whereas not pipelining makes the cause of
        the error message obvious.

@@

In the case where the \daemon{cleanup} line is still pending during the
TIMEOUT action the algorithm updates a global cache of queueids, adding the
queueid and the timestamp from the timeout line.  When the next
\daemon{cleanup} line is parsed for that queueid the cache will be checked,
and the line will be deemed part of the discarded mail and discarded if it
meets the following requirements:

@@

The next \daemon{cleanup} line must meet the criteria above for it to be
discarded because not every connection where a timeout occurs will have a
\daemon{cleanup} line logged for it; if the algorithm blindly discarded the
next \daemon{cleanup} line after a TIMEOUT it would in some cases be
mistaken.  Whether or not the next \daemon{cleanup} line is discarded the
queueid will be removed from the cache of timeout queueids when the next
\daemon{pickup} line containing that queueid is parsed.

During the TIMEOUT action if the mail lacks a \daemon{cleanup} line, the
queueid and other data about the mail, particularly the timestamp of the
last log line, are added to a global cache of timeout queueids.  This cache
is subsequently consulted during the MAIL\_PICKED\_FOR\_DELIVERY action,
and if the queueid is found in the cache and the \daemon{cleanup} line
passes the checks listed above the line will be discarded.

@@

Occasionally the \daemon{pickup} line logged when mail is submitted locally
via sendmail appears later in the log file than the \daemon{cleanup} line
for that mail.  This seems to occur during periods of particularly heavy
load, so is most likely due to process scheduling vagaries.  Normally if
the queueid given in the \daemon{pickup} line exists a warning is generated
by the \daemon{pickup} action, but if the following conditions are met it
is assumed that the lines are out of order:

@@

    \item There is less than a five second difference between the
        timestamps of the \daemon{cleanup} and \daemon{pickup} lines.

@@

As always with heuristics there may be circumstances in which these
heuristics match incorrectly,  but none have been identified so far.

@@

    \item Postfix is stopped or its configuration is reloaded.  When this
        happens all \daemon{smtpd} processes exit, and all entries in the
        connections state table must be cleaned up, entered in the database
        and deleted.  The queueid state table is untouched.

    \item Postfix sometimes forcibly kills a \daemon{smtpd} XXX FIGURE OUT
        WHY, so the active connection must be cleaned up, entered in the
        database and deleted from the connections state table.

    \item Occasionally a \daemon{smtpd} will die or exit uncleanly, so the
        active connection must be cleaned up, entered in the database and
        deleted from the connections state table.

@@

The above descriptions appear to cover all situations identified thus far
where a \daemon{smtpd} suddenly stops logging.  In addition to removing an
active connection the last accepted mail may need to be discarded, as
detailed in section~\refwithpage{timeouts-during-data-phase}.

@@

Occasionally a log file will have out of order log lines which cannot be
dealt with by the techniques described in sections~\refwithpage{tracking
re-injected mail},~\refwithpage{discarding cleanup
lines}~or~\refwithpage{pickup logging after cleanup}.  In the 93 log files
used for testing this occurs only five times in 60,721,709 log lines, but
for completeness of the algorithm it should be dealt with.  The five
occurrences in the test log files have the same characteristics: the
\daemon{local} log line showing delivery to a local mailbox occurs after
the \daemon{qmgr} log line showing removal of the mail from the queue
because delivery is completed.  This causes problems: the mail is not
complete, so entry into the database fails; a new mail is created when the
\daemon{local} line is parsed and remains in the state tables; four
warnings are issued per pair of out of order log lines.

The solution to this problem is to examine the list of programs which have
logged messages for each mail, comparing the list against a table of
known-good combinations of programs (this check is performed during the
COMMIT action).  If the mail's combination is found in the valid list the
mail can be entered in the database; if the combination is not found entry
must be postponed and the mail flagged for later entry.  The
SAVE\_BY\_QUEUEID action checks for the flag and retries entry if it's
found; if the additional log lines have caused the mail to reach a valid
combination entry will proceed, otherwise it must be postponed once more.

The list of valid combinations is explained below.  Every mail will
additionally have log entries from \daemon{cleanup} and \daemon{qmgr}; any
mail may also have log entries from \daemon{bounce}, \daemon{postsuper}, or
both.

@@

    \item [\daemon{local}:] Local delivery of a bounce notification, or
        local delivery of a forwarded or tracked mail.

@@

    \item [\daemon{local}, \daemon{smtpd}:] Mail accepted from a remote
        client, local delivery only.

    \item [\daemon{pickup}, \daemon{smtp}:] Locally submitted mail, remote
        delivery only.

    \item [\daemon{smtp}:] Remote delivery of forwarded or tracked mail, or
        a bounce notification.

@@

This check applies to accepted mails only, not to rejected mails.

@@

This section has presented the parsing algorithm, starting with a high
level overview of the basic algorithm, and the first group of difficulties
thwarted.  The revised algorithm was detailed, preceded by a flow chart
depicting the paths a mail may take through the algorithm, and followed by
the actions which are available to rules were described in full.  The
section concludes with the remaining complications discovered and overcome
in the process of refining and completing the algorithm.  Detecting,
diagnosing and defeating complications forms the largest single portion of
this document, mirroring the development of the parser,  The complications
are described in the order they were overcome, with subsequent problems
affecting fewer mails (often by an order or magnitude), though the time
required to solve problems increased with each successive problem.

@@

Coverage of mails is much more difficult to determine accurately than
coverage of lines.  The parser can dump its state tables in a human
readable form; examining these tables with reference to the log files is
the best way to detect mails which were not handled properly (many of the
complications discussed in section~\refwithpage{additional complications}
were detected in this way).  The parser issues warnings when it detects any
errors, some of which may alert the user to a problem, e.g.\ when a queueid
is reused before the previous mail is fully dealt with, when a queueid or
\pid{} is not found,\footnote{There will often be warnings about a missing
queueid or \pid{} in the first few hundred or thousand log lines because the
earlier log lines for those connections or mails are in the previous log
file; loading the saved state from the previous log file will solve this
problem.} or when there are problems tracking a child mail (see
section~\refwithpage{tracking re-injected mail}).  There should be few or
no warnings, and the state table should only contain entries for mails
which had yet to be delivered when the log files ended, or started before
the log files began.

@@

The dips at log files 22 and 62--68 correspond to peaks in log file size in
graph~\refwithpage{execution time vs file size vs number of lines graph},
and peaks in graph~\refwithpage{execution time vs file size vs number lines
factor} (where a peak means that more lines are processed per second, i.e.\
performance is better).  The explanation for this took some time to arrive
at, but it turns out to be reasonably simple.  The large log files in
question were caused by a mail forwarding loop, where the distribution of
log lines is quite different to normal, resulting in different performance
characteristics.

@@

% The contents of the glossary.
\glossary{name={SQLite3},description={
    \textit{SQLite is a small C library that implements a self-contained,
    embeddable, zero-configuration SQL database engine.\/} SQLite3 is an
    \SQL{} implementation focusing on correctness, simplicity and speed.
    Unlike other \SQL{} implementations it does not require a separate
    server process, greatly simplifying deployment of an application
    utilising it.  More details can be found at~\cite{sqlite-features} or
    \url{http://www.sqlite.org/}.
}}

@@

\glossary{name={queueid},description={
    Each mail in Postfix's queue is assigned a queueid to uniquely identify
    it.  Queueids are assigned from a limited pool, so although they are
    guaranteed to be unique for the lifetime of the mail, given sufficient
    time they will be reused.
}}
